---
title: "CptS 475 HW2"
date: "September 4th, 2023"
author: Chase Conaway
pdf_document: default
---
# Homework 2: R Basics and Exploratory Data Analysis

## Question 1: Red wine quality

Let's load the library ggplot2 first.
```{r}
# load the ggplot2 library
library(ggplot2)
```

Next, we can load the dataset winequality.csv into an object we can use for the whole question.

### Part a: Loading the winequality csv
```{r}
redwine <- read.csv('./winequality-red.csv')
head(redwine)
str(redwine)
```

We load the table and see that our data looks good. Nice.

### Part b: Mean quality and alcohol levels

```{r}
meanQuality <- mean(redwine$quality)
meanAlcContent <- mean(redwine$alcohol)
meanQuality
meanAlcContent
```

### Part c: Scatterplot: density vs residual sugar

```{r}
plot(log(redwine$density), log(redwine$residual_sugar), xlab="Density", ylab="Residual Sugar", 
            main="Density vs. Residual Sugar")
```

We plotted the log values of both the density and residual_sugar. The log values help the relationship become a little clearer.

We see that a higher density is somewhat correlated with a higher residual sugar content, which, makes sense. If there is more sugar in the wine, we are raising the density of it by having more stuff in it.

### Part d: High and Medium Alcohol

```{r}
alcoholLevels <- c("Medium", "High")

# cut the alc content into levels based on if they are > 11 or not and add to the dataframe
redwine$ALevels <- as.factors(cut(redwine$alcohol, breaks=c(0, 11, 1000), labels=alcoholLevels))
redwine$StoCl <- redwine$sulphates / redwine$chlorides
head(redwine)

ggplot(data=redwine, aes(x=ALevels, y=StoCl, fill=ALevels)) + 
    geom_boxplot(alpha=0.3) + 
    theme_minimal() + 
    labs(x="Alcohol Content", y="Sulphate to Chlorides ratio", title="Alcohol Content Boxplot", 
        caption="Red wine data") + 
    theme(plot.title = element_text(hjust = 0.5), 
        axis.text.x = element_text(angle = 45, hjust = 0.8))

sum(redwine$ALevels == "High")
```

### Part e: Fixed acidity histogram

```{r}
highA <- hist(redwine$fixed_acidity[redwine$ALevels == "High"], breaks=50, plot=FALSE)
mediumA <- hist(redwine$fixed_acidity[redwine$ALevels == "Medium"], breaks=50, plot=FALSE)
plot(mediumA, col=rgb(0,1,0, 1/4), xlab="Fixed Acidity", main="Fixed acidity frequencies")
plot(highA, col=rgb(0,0,1,1/4), add=T)
legend("topright", c("High", "Medium"), col=c(rgb(0,0,1,1/4), rgb(0,1,0,1/4)), lwd=10)
```

### Part f: Two more plots

First plot: volatile acidity vs total sulfur dioxide 
I am unsure what this is in wine, but I hypothesize there will be almost no correlation
```{r}
reg <- lm(redwine$volatile_acidity ~ redwine$total_sulfur_dioxide)
reg
ggplot(redwine, aes(x=log(volatile_acidity), y=log(total_sulfur_dioxide), col=volatile_acidity)) + 
geom_point() +
geom_abline(intercept=0.5085, slope=0.004163) +
scale_colour_gradient(low="#eb4034", high="#4430b8")
```

I went ahead and added the line to the plot, but the slope is almost 0, which indicates 0 relationship.

Second plot: ph and total sulfur dioxide
I predict that if SO2 is acidic, the wine will have a higher pH, and if it is basic, it will be lower.
```{r}
reg <- lm(redwine$pH ~ redwine$total_sulfur_dioxide)
coeff <- coefficients(reg)
ggplot(redwine, aes(x=pH, y=total_sulfur_dioxide, col=pH)) +
geom_point() + 
geom_abline(intercept=coeff[1], slope=coeff[2], size=2) +
scale_colour_gradient(low="#eb4034", high="#4430b8")
```

My hypothesis was wrong, there is no correlation, once again.

## Question 2: Forest fires

First let's get our csv into an objext we can use

```{r}
forestfires <- read.csv("./forestfires.csv")
head(forestfires)
str(forestfires)
```

### Part a: Qualitative and quantitative predictors

Month and day are qualitative. They let us know the month and day the fire started. Everything else is qualitative in nature, describing conditions of the environment numerically.

### Part b: Range, mean, and standard deviation of each quantitative predictor

```{r}
# Convert DC to numeric type
forestfires$DC <- as.numeric(forestfires$DC)
for (column in colnames(forestfires)) {
    print(column)
    if (is.numeric(forestfires[[column]])) {
        cat("Mean: ", as.character(mean(forestfires[[column]], na.rm=TRUE)), "\n")
        cat("Standard deviation: ", as.character(sd(forestfires[[column]], na.rm=TRUE)), "\n")
        cat("Range: ", as.character(range(forestfires[[column]], na.rm=TRUE)), "\n")
    }
}

numOfFiresPerMonth <- table(forestfires$month)
cat("Month with the max number of fires:", names(which.max(numOfFiresPerMonth)), "with", max(numOfFiresPerMonth[2]), "fires\n")
```